DRAFT: The algebraic hierarchy in type theory using type classes

### 2010-02-26 processed.tex
Major rewrite.

\documentclass[a4paper,10pt, runningheads]{llncs}
\usepackage{url}
\usepackage{bbm}
\usepackage{ucs}
% \usepackage[utf8x,utf8]{inputenc}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
%% ODER: format ==         = "\mathrel{==}"
%% ODER: format /=         = "\neq "
%
%
\makeatletter
\@ifundefined{lhs2tex.lhs2tex.sty.read}%
  {\@namedef{lhs2tex.lhs2tex.sty.read}{}%
   \newcommand\SkipToFmtEnd{}%
   \newcommand\EndFmtInput{}%
   \long\def\SkipToFmtEnd#1\EndFmtInput{}%
  }\SkipToFmtEnd

@@

\arrayhs%i.e. no page breaks in code blocks
\divide\abovedisplayskip 4
\divide\belowdisplayskip 4
\begin{document}
\title{The algebraic hierarchy in type theory using type classes}
\author{Bas Spitters \and Eelis van der Weegen}
\institute{Radboud University Nijmegen}
\maketitle
\begin{abstract}
We propose generic design patterns to organize algebraic structures using type classes inside the
Coq system. We support multiple inheritance, sharing of notations and theories, and automated
structure inference. A library for many-sorted universal algebra abstracts from the specific
algebraic structures and aids the organization of the library. Category theory provides a further
useful abstraction. We provide abstract
interfaces for $\N$, $\Z$, $\Q$ as the initial semiring, the initial ring and a field of
fractions. Similarly, we provide an interface for polynomials over a ring $R$ using $R$-algebras.
We show that the standard implementations of these are in fact instances of these type classes.
We report on the efficiency penality of these abstractions.

@@

\subsection{Type classes}
Implementation in Ltac, Notation etc
Coercions

@@

We \emph{could} become over-eager and decide to unbundle the equivalence proof term as well:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{Record}\;\Conid{Monoid}\;(\Conid{A:}\;\Conid{Type})\;(\Conid{R:}\;\Conid{A}\;\rightarrow\;\Conid{A}\;\rightarrow\;\Conid{Prop})\;{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}(\Varid{e:}\;\Conid{Equivalence}\;\mathbb{R})\;\mathbin{:}\;\Conid{Type}\;\Conid{:=}\;\Varid{...}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
However, this way if we happen to have two proofs \ensuremath{\Varid{p}} and \ensuremath{\Varid{q}} that a certain relation is an
equivalence, we end up with two non-convertible ways to express that the relation (together with
some unit and binary operation) form a Monoid: 'Monoid A R p ...' and 'Monoid A R q ...'. We must
either try to deal with this (by proving that Monoid proofs with different Equivalence proofs are
interchangable, and hope for automation, such as type class instance resolution, to use them
interchangably), or take great care to ensure that \ensuremath{\Conid{Equivalence}} proof terms are always shared,
essentially elevating them to `values', which goes against the principle of proof
irrelevance.\marginpar{What if we add proof irrelevance?}

Neither of these options are very appealing. We really should just make \ensuremath{\Conid{Monoid}} take as parameters
only the operations/relations. If we treat the monoid's unit and binary operation the same way, this
means that \ensuremath{\Conid{Monoid}} ends up in \ensuremath{\Conid{Prop}}, and that its instances can be left opaque, which is really
how things ought to be.

Another benefit of this approach is that it works very smoothly when these records are in fact type
classes. When relations and operations are bundled and do \emph{not} appear as parameters of the
structure record/class, expressing as a type class constraint that some collection of relations and
operations posess the structure seems to require manifest fields; see Section~\ref{manifest}. With
our unbundled approach, on the other hand, such type class constraints are easily expressed without
any extensions as simple type inhabitation queries
% (e.g. ``hey Coq, find me a proof of [Monoid some_rel some_op some_unit]'')
of types living in Prop, so that we don't even care what instances are found (thus embracing proof
irrelevance).

We remark that leaving the equivalence proofs bundled means that when one has a proof that something
is a semiring, one has two proofs (via the two monoids) that the equivalence relation is an
equivalence. Fortunately, having redundant proofs like this is entirely harmless when one's terms
never refer (directly or indirectly) to proofs, which is the case when the relations and operations
they are composed of are left unbundled. Here, too, we enjoy the warm embrace of proof irrelevance.

From the discussion so far, we derive some principles:
\begin{enumerate}
 \item structural concepts (like Reflexive, Setoid, Monoid) should have their structural components
(i.e. carriers, relations, and operations) as parameters, and nothing else;
 \item (as a consequence) these concepts should be defined as predicates (which includes records and
classes living in Prop), and their proofs may be kept opaque;
 \item (as a second consequence) algebraic expressions need/should never refer (directly or
indirectly) to proofs.
\end{enumerate}
% \setcounter{enumi}{3}

@@

Suppose we have defined \ensuremath{\Conid{SemiGroup}} in line with the aforementioned principles:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathbf{Record}\;\Conid{SemiGroup}\;(\Conid{A:}\;\Conid{Type})\;(\Varid{eq:}\;\Conid{A}\;\rightarrow\;\Conid{A}\;\rightarrow\;\Conid{Prop})\;(\Varid{op:}\;\Conid{A}\;\rightarrow\;\Conid{A}\;\rightarrow\;\Conid{A})\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}\;\Varid{...}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Then our starting point is the obvious but awful:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathbf{Lemma}\;\Varid{assoc1}\;\Conid{A}\;\Varid{eq}\;\Varid{op}\;(\Conid{P:}\;\Conid{SemiGroup}\;\Conid{A}\;\Varid{eq}\;\Varid{op})\;\Varid{x}\;\Varid{y}\;\Varid{z:}\;{}\<[E]%
\\
\>[3]{}\hsindent{17}{}\<[20]%
\>[20]{}\Varid{eq}\;(\Varid{op}\;(\Varid{op}\;\Varid{x}\;\Varid{y})\;\Varid{z})\;(\Varid{op}\;(\Varid{op}\;\Varid{z}\;\Varid{y})\;\Varid{x})\;.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
As a first cleaning measure, we might hope to introduce some notations to let us use infix notation
for the semigroup operator (say, \ensuremath{\Varid{\&}}) and for equality. Unfortunately, at this point the names \ensuremath{\Varid{eq}}
and \ensuremath{\Varid{op}} are just local names and there is no way to predeclare that they are to be associated with
any notation. What we really need are canonical names for these notions, so that we may fashion
notations for them.

@@

These single-field type classes are given special treatment in that they are not translated into
records the way classes with any other number of fields are, but are instead turned into ordinary
definitions. The effect of this is that \ensuremath{\Conid{Equiv}\;\Conid{A}} immediately reduces to \ensuremath{\Varid{relation}\;\Conid{A}}, and that the
equiv `projection' is the identity function whose argument is inferred using instance resolution.
Thus, for a very modest cost in indirection, we gain the ability to use canonical names, and with
it, notations:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathbf{Record}\;\Conid{SemiGroup}\;(\Conid{A:}\;\Conid{Type})\;(\Varid{eq:}\;\Conid{Equiv}\;\Conid{A})\;{}\<[E]%
\\
\>[3]{}\hsindent{27}{}\<[30]%
\>[30]{}(\Varid{op:}\;\Conid{SemiGroupOp}\;\Conid{A})\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}\;\Varid{...}{}\<[E]%
\\
\>[3]{}\mathbf{Lemma}\;\Varid{assoc1}\;\Conid{A}\;\Varid{eq}\;\Varid{op}\;(\Conid{P:}\;\Conid{SemiGroup}\;\Conid{A}\;\Varid{eq}\;\Varid{op})\;\Varid{x}\;\Varid{y}\;\Varid{z:}\;{}\<[E]%
\\
\>[3]{}\hsindent{18}{}\<[21]%
\>[21]{}\Varid{x}\;\Varid{\&}\;\Varid{y}\;\Varid{\&}\;\Varid{z}\;\mathrel{=}\;\Varid{z}\;\Varid{\&}\;\Varid{y}\;\Varid{\&}\;\Varid{x.}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
We call these type classes \emph{operational type classes}. We note that operational type classes
allow us to avoid Coq's notation \ensuremath{\mathbf{scope}} mechanism.

It is important to point out that although the expression now \emph{looks} like it is bound to some
specific semigroup structure (which with traditional approaches would imply projections from bundles
and/or reference to proofs), it is really only referring near-directly to the actual operations
involved, with the semigroup proof existing only as opaque `knowledge about the operations' which we
may use in the proof. This lack of projections keeps our terms small and independent, and keeps
rewriting simple and sane.

On the other hand, it \emph{does} mean that all the carriers/relations/operations end up in the
context. The context for \ensuremath{\Varid{assoc1}} looks like:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\Conid{A:}\;\Conid{Type}{}\<[E]%
\\
\>[3]{}\Varid{eq:}\;\Conid{Equiv}\;\Conid{A}{}\<[E]%
\\
\>[3]{}\Varid{op:}\;\Conid{SemiGroupOp}\;\Conid{A}{}\<[E]%
\\
\>[3]{}\Conid{P:}\;\Conid{SemiGroup}\;\Conid{A}\;\Varid{eq}\;\Varid{op}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

While we are not particularly worried about overly large contexts (we feel that this will most
likely not be problematic for any but the most complex formalizations), having to \emph{declare}
these entities (like \ensuremath{\Varid{eq}} and \ensuremath{\Varid{op}} for \ensuremath{\Varid{assoc1}}) \emph{is} a bit of a chore. Fortunately, Coq
provides a feature called implicit generalization, which is exactly what we need.
\marginpar{describe how it works} With it, we can write \ensuremath{\Varid{assoc1}} as follows:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathbf{Lemma}\;\Varid{assoc1}\hspace*{0mm}`(\Conid{SemiGroup}\;\Conid{A})\;\Varid{x}\;\Varid{y}\;\Varid{z:}\;\Varid{x}\;\Varid{\&}\;\Varid{y}\;\Varid{\&}\;\Varid{z}\;\mathrel{=}\;\Varid{z}\;\Varid{\&}\;\Varid{y}\;\Varid{\&}\;\Varid{x.}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Thus, we have reached the mathematical ideal we aimed for.

@@

\section{Subset types}
BEGIN SPECULATIVE

A record type is \marginpar{mostly?} syntactic sugar for a $\Sigma$-type. In the presence of a
squash operator, \ensuremath{[\mskip1.5mu \mskip1.5mu]\;\Conid{:Type->Prop}}, we may define a subset type as \ensuremath{\lambda \;\hspace*{0mm}\{\Conid{A\char92 mid}\;\Conid{B\char92 }\mskip1.5mu\}\;\Varid{=Sig}\;\Conid{A.}\;[\mskip1.5mu \Conid{B}\mskip1.5mu]}. This is
pursued by Barras~\cite{Barras:subset,Werner}. The program machinery fakes subset types in the
present Coq implementation. Proof irrelevanve in Coq would allow a faithful
implementation.

Squash is a monad and allows many constructions: subsets, unions, images, heterogeneous equality,
\ldots.

Question: Would the meta-theory simplify if we would just add a conversion rule $x\triangleright
\epsilon$ for $x:Prop$? We could add a term constructor $[a]$ which will have type $[A]$. There
seems to be no reason to connect this to \ensuremath{\Conid{Prop}} or impredicativity.

END SPECULATIVE

We add another principle:
\begin{itemize}\setcounter{enumi}{3}
 \item Working hypothesis: `every record can be reordered as a subset'. Consequently, we
rewrite this subset \ensuremath{\lambda \;\hspace*{0mm}\{\Conid{A\char92 mid}\;\Conid{B\char92 }\mskip1.5mu\}} as \ensuremath{\mathbf{Record}\;\Varid{r}\;(\Varid{a:A})\;\Conid{:=}\;\hspace*{0mm}\{\Varid{b:B}\mskip1.5mu\}}.
\end{itemize}

\section{Algebraic hierarchy using type classes}\label{classes}
\paragraph{Hierarchy.}\label{hierarchy}
In our formalization of algebraic structures we use the structure as (implicit) parameters, but
pack the proofs. This can be nicely organized by using canonical names. For instance, we define:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{Class}\;\Conid{SemiRing}\;\Conid{A}\;\hspace*{0mm}\{\Varid{e:}\;\Conid{Equiv}\;\Conid{A}\mskip1.5mu\}\;\hspace*{0mm}\{\Varid{plus:}\;\Conid{RingPlus}\;\Conid{A}\mskip1.5mu\}\;\hspace*{0mm}\{\Varid{mult:}\;\Conid{RingMult}\;\Conid{A}\mskip1.5mu\}\;{}\<[E]%
\\
\>[B]{}\hsindent{18}{}\<[18]%
\>[18]{}\hspace*{0mm}\{\Varid{zero:}\;\Conid{RingZero}\;\Conid{A}\mskip1.5mu\}\;\hspace*{0mm}\{\Varid{one:}\;\Conid{RingOne}\;\Conid{A}\mskip1.5mu\}\;\Conid{:Prop}\;\Conid{:=}\;{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\hspace*{0mm}\{\Varid{semiring\char95 mult\char95 monoid:>}\;\Conid{Monoid}\;\Conid{A}\;(\Varid{op:=mult})\;(\Varid{unit:=one}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{};\Varid{semiring\char95 plus\char95 monoid:>}\;\Conid{Monoid}\;\Conid{A}\;(\Varid{op:=plus})\;(\Varid{unit:=zero}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{};\Varid{semiring\char95 plus\char95 comm:>}\;\Conid{Commutative}\;\Varid{plus}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{};\Varid{semiring\char95 mult\char95 comm:>}\;\Conid{Commutative}\;\Varid{mult}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{};\Varid{semiring\char95 distr:>}\;\Conid{Distribute}\;\Varid{mult}\;\Varid{plus}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{};\Varid{mult\char95 0\char95 l:}\;\forall\;\Varid{x},\Varid{0}\;\Varid{*}\;\Varid{x}\;= \;\Varid{0}\mskip1.5mu\}\;.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
where e.g.\ \ensuremath{\Conid{RingPlus}} is the operation class for the semigroup operator.
The notation \ensuremath{\Conid{:>}} makes \ensuremath{\Varid{semiring\char95 mult\char95 monoid}} a coercion: when asked for a monoid it suffices to
provide a ring.

\emph{Structure-as-parameters} helps setoid-rewriting: type class resolution
can find the equivalence relation in the context.
% A similar style should be possible for, say, the |Ring| tactic, instead of
% declaring the ring structure by a separate command, we would rely on type class resolution to find
% it.
We note that \ensuremath{\Varid{op}} does not depend on the proof that \ensuremath{\Varid{e}} is an equivalence. We use Coq's implicit
quantification (\ensuremath{\hspace*{0mm}`\hspace*{0mm}\{\mskip1.5mu\}}) to avoid having to write all the parameters when \emph{stating} a theorem
and Coq's maximally inserted implicit arguments to find the parameters when \emph{applying} a
theorem. Both features are new in Coq and stem from the type class implementation.

We mention the trade-off between bigger contexts versus bigger terms. Our contexts are bigger than
those of telescopes or packed classes. In our experience, this has been relatively
harmless\footnote{Coq's data structure for contexts is not very efficient however, we believe this
may be a bottleneck for huge developments. It seems that the data structure chosen
in~\cite{asperti2009compact} will behave better.}%
: most
terms in the context are there to support canonical names. Bigger terms
\emph{do} cause problems: 1. when proofs are part of mathematical objects we need to share these
proofs to allow rewriting. Moreover, it prohibits Opaque proofs and `proof irrelevance'. 2. The
projection paths may not be canonical.

@@

We provide the category of setoids, the category of categories, the dual of a category, products,
and hence co-products, the category of $\Sigma$-algebras.

>>>>

\documentclass[a4paper,10pt, runningheads]{llncs}
\usepackage{url}
\usepackage{bbm}
\usepackage{ucs}
\usepackage{amsmath}
% \usepackage[utf8x,utf8]{inputenc}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
%% ODER: format ==         = "\mathrel{==}"
%% ODER: format /=         = "\neq "
%
%
\makeatletter
\@ifundefined{lhs2tex.lhs2tex.sty.read}%
  {\@namedef{lhs2tex.lhs2tex.sty.read}{}%
   \newcommand\SkipToFmtEnd{}%
   \newcommand\EndFmtInput{}%
   \long\def\SkipToFmtEnd#1\EndFmtInput{}%
  }\SkipToFmtEnd

@@

%%%format R = "\mathbb{R}"

\arrayhs%i.e. no page breaks in code blocks
\divide\abovedisplayskip 4
\divide\belowdisplayskip 4
\begin{document}
\title{DRAFT: The algebraic hierarchy in type theory using type classes}
\author{Bas Spitters \and Eelis van der Weegen}
\institute{Radboud University Nijmegen}
\maketitle
\begin{abstract}
We propose generic design patterns to organize algebraic structures using type classes inside the
Coq system. We support multiple inheritance, sharing of notations and theories, and automated
structure inference. A library for many-sorted universal algebra abstracts from the specific
algebraic structures and aids the organization of the library. Category theory provides a further
useful abstraction. We provide abstract
interfaces for $\N$, $\Z$, $\Q$ as the initial semiring, the initial ring and a field of
fractions. Similarly, we provide an interface for polynomials over a ring $R$ using $R$-algebras.
We show that the standard implementations of these are in fact instances of these type classes.
We report on the efficiency penality of these abstractions.

@@

Future versions of Coq are planned to support proof irrelevance: every \ensuremath{\Varid{p},\Varid{q:P:Prop}} are
convertible; see section~\cite{explit}.

\subsection{Type classes}
Implementation in Ltac, Notation etc\\
Coercions\\
Unification hints

@@

% We \emph{could} become over-eager and decide to unbundle the equivalence proof term as well:
% \begin{code}
% Record Monoid (A: Type) (R: A → A → Prop)
%   (e: Equivalence R) : Type := ...
% \end{code}
% However, this way if we happen to have two proofs |p| and |q| that a certain relation is an
% equivalence, we end up with two non-convertible ways to express that the relation (together with
% some unit and binary operation) form a Monoid: 'Monoid A R p ...' and 'Monoid A R q ...'. We must
% either try to deal with this (by proving that Monoid proofs with different Equivalence proofs are
% interchangable, and hope for automation, such as type class instance resolution, to use them
% interchangably), or take great care to ensure that |Equivalence| proof terms are always shared,
% essentially elevating them to `values', which goes against the principle of proof
% irrelevance.\marginpar{What if we add proof irrelevance?}
% BOTTOMLINE: Bundle as much as we can!
% 
% Neither of these options are very appealing. We really should just make |Monoid| take as
% parameters
% only the operations/relations. If we treat the monoid's unit and binary operation the same way,
% this
% means that |Monoid| ends up in |Prop|, and that its instances can be left opaque, which is really
% how things ought to be.

Another benefit of our approach is that it works very smoothly when these records are in fact type
classes. When relations and operations are bundled and do \emph{not} appear as parameters of the
structure record/class, to express that some collection of relations and
operations posess the structure as a type class constraint seems to require manifest fields; see
Section~\ref{manifest}. With
our unbundled approach, on the other hand, such type class constraints are easily expressed without
any extensions as type inhabitation queries
% (e.g. ``hey Coq, find me a proof of [Monoid some_rel some_op some_unit]'')
of types living in Prop, so that we don't even care what instances are found (thus embracing proof
irrelevance).

Leaving the equivalence proofs bundled means that when one has a proof that a structure
is a semiring, one has two proofs (via the two monoids) that the equivalence relation is an
equivalence. Fortunately, this redundancy is entirely harmless since our terms
never refer (directly or indirectly) to proofs, which is the case when the relations and operations
they are composed from are left unbundled. Here, too, we enjoy the warm embrace of proof
irrelevance.

From the discussion so far, we derive some principles:
\begin{enumerate}
 \item structural concepts (like Reflexive, Setoid, Monoid) should have their structural components
(i.e. carriers, relations, and operations) as parameters, and nothing else;
 \item (as a consequence) these concepts should be defined as predicates (which includes records and
classes living in Prop), and their proofs may be kept opaque;
 \item (as a second consequence) algebraic expressions need/should never refer (directly or
indirectly) to proofs about the algebraic structure.\footnote{The reciprocal in a field
famously does refer to a proof that the element is non-zero, but not to a proof that the operations
indeed form a field.}
\end{enumerate}
% \setcounter{enumi}{3}

@@

Suppose we have defined \ensuremath{\Conid{SemiGroup}} in line with the aforementioned principles:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{Record}\;\Conid{SemiGroup}\;(\Conid{A:}\;\Conid{Type})\;(\Varid{eq:}\;\Conid{A}\;\rightarrow\;\Conid{A}\;\rightarrow\;\Conid{Prop})\;{}\<[E]%
\\
\>[B]{}(\Varid{op:}\;\Conid{A}\;\rightarrow\;\Conid{A}\;\rightarrow\;\Conid{A})\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}\;\Varid{...}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Then our starting point is the obvious but awful:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{Lemma}\;\Varid{perm}\;\Conid{A}\;\Varid{eq}\;\Varid{op}\;(\Conid{P:}\;\Conid{SemiGroup}\;\Conid{A}\;\Varid{eq}\;\Varid{op})\;\Varid{x}\;\Varid{y}\;\Varid{z:}\;{}\<[E]%
\\
\>[B]{}\hsindent{20}{}\<[20]%
\>[20]{}\Varid{eq}\;(\Varid{op}\;(\Varid{op}\;\Varid{x}\;\Varid{y})\;\Varid{z})\;(\Varid{op}\;(\Varid{op}\;\Varid{z}\;\Varid{y})\;\Varid{x})\;.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
As a first cleaning measure, we might hope to introduce some notations to let us use infix notation
for the semigroup operator (say, \ensuremath{\Varid{\&}}) and for equality. Unfortunately, at this point the names \ensuremath{\Varid{eq}}
and \ensuremath{\Varid{op}} are just local names and there is no way to predeclare that they are to be associated with
any notation. What we really need are canonical names for these notions, so that we may fashion
notations for them.

@@

These single-field type classes are given special treatment in that they are not translated into
records the way classes with any other number of fields are, but are instead turned into ordinary
definitions. The effect of this is that \ensuremath{\Conid{Equiv}\;\Conid{A}} immediately reduces to \ensuremath{\Varid{relation}\;\Conid{A}}, and that the
equiv `projection' is the identity function whose argument is inferred using instance resolution.
Thus, for a very modest cost in indirection, we gain the ability to use canonical names, and with
it, notations:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{Record}\;\Conid{SemiGroup}\;(\Conid{A:}\;\Conid{Type})\;(\Varid{eq:}\;\Conid{Equiv}\;\Conid{A})\;{}\<[E]%
\\
\>[B]{}\hsindent{18}{}\<[18]%
\>[18]{}(\Varid{op:}\;\Conid{SemiGroupOp}\;\Conid{A})\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}\;\Varid{...}{}\<[E]%
\\
\>[B]{}\mathbf{Lemma}\;\Varid{perm}\;\Conid{A}\;\Varid{eq}\;\Varid{op}\;(\Conid{P:}\;\Conid{SemiGroup}\;\Conid{A}\;\Varid{eq}\;\Varid{op})\;\Varid{x}\;\Varid{y}\;\Varid{z:}\;{}\<[E]%
\\
\>[B]{}\hsindent{18}{}\<[18]%
\>[18]{}\Varid{x}\;\Varid{\&}\;\Varid{y}\;\Varid{\&}\;\Varid{z}\;\mathrel{=}\;\Varid{z}\;\Varid{\&}\;\Varid{y}\;\Varid{\&}\;\Varid{x.}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
We call these type classes \emph{operational type classes}. We note that operational type classes
allow us to avoid Coq's notation \ensuremath{\mathbf{scope}} mechanism.

It is important to point out that although the expression now \emph{looks} like it is bound to some
specific semigroup structure (which with traditional approaches would imply projections from bundles
and/or reference to proofs), it is really only referring near-directly to the actual operations
involved, with the semigroup proof existing only as opaque `knowledge about the operations' which we
may use in the proof. This lack of projections keeps our terms small and independent, and keeps
rewriting simple and sane; see Section~\ref{canonical}.

On the other hand, it \emph{does} mean that all the carriers/relations/operations end up in the
context. The context for \ensuremath{\Varid{perm}} looks like:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\Conid{A:}\;\Conid{Type}{}\<[E]%
\\
\>[3]{}\Varid{eq:}\;\Conid{Equiv}\;\Conid{A}{}\<[E]%
\\
\>[3]{}\Varid{op:}\;\Conid{SemiGroupOp}\;\Conid{A}{}\<[E]%
\\
\>[3]{}\Conid{P:}\;\Conid{SemiGroup}\;\Conid{A}\;\Varid{eq}\;\Varid{op}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

While we are not particularly worried about overly large contexts (we feel that this will most
likely not be problematic for any but the most complex formalizations), having to \emph{declare}
these entities (like \ensuremath{\Varid{eq}} and \ensuremath{\Varid{op}} for \ensuremath{\Varid{perm}}) \emph{is} a bit of a chore. Fortunately, Coq
provides a feature called implicit generalization, which is exactly what we need.
With it, we can write \ensuremath{\Varid{perm}} as follows:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathbf{Lemma}\;\Varid{perm}\hspace*{0mm}`(\Conid{SemiGroup}\;\Conid{A})\;\Varid{x}\;\Varid{y}\;\Varid{z:}\;\Varid{x}\;\Varid{\&}\;\Varid{y}\;\Varid{\&}\;\Varid{z}\;\mathrel{=}\;\Varid{z}\;\Varid{\&}\;\Varid{y}\;\Varid{\&}\;\Varid{x.}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Thus, we have reached the mathematical ideal we aimed for.

@@

\section{Putting it all together}\label{classes}
Combining the ideas in the previous sections we end up with a formalization of algebraic
structures where we use the structure as (implicit) parameters, but
pack the proofs. This can be nicely organized by using canonical names. For instance, we define:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{Class}\;\Conid{SemiRing}\;\Conid{A}\;\hspace*{0mm}\{\Varid{e:}\;\Conid{Equiv}\;\Conid{A}\mskip1.5mu\}\;\hspace*{0mm}\{\Varid{plus:}\;\Conid{RingPlus}\;\Conid{A}\mskip1.5mu\}\;\hspace*{0mm}\{\Varid{mult:}\;\Conid{RingMult}\;\Conid{A}\mskip1.5mu\}\;{}\<[E]%
\\
\>[B]{}\hsindent{18}{}\<[18]%
\>[18]{}\hspace*{0mm}\{\Varid{zero:}\;\Conid{RingZero}\;\Conid{A}\mskip1.5mu\}\;\hspace*{0mm}\{\Varid{one:}\;\Conid{RingOne}\;\Conid{A}\mskip1.5mu\}\;\Conid{:Prop}\;\Conid{:=}\;{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\hspace*{0mm}\{\Varid{semiring\char95 mult\char95 monoid:>}\;\Conid{Monoid}\;\Conid{A}\;(\Varid{op:=mult})\;(\Varid{unit:=one}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{};\Varid{semiring\char95 plus\char95 monoid:>}\;\Conid{Monoid}\;\Conid{A}\;(\Varid{op:=plus})\;(\Varid{unit:=zero}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{};\Varid{semiring\char95 plus\char95 comm:>}\;\Conid{Commutative}\;\Varid{plus}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{};\Varid{semiring\char95 mult\char95 comm:>}\;\Conid{Commutative}\;\Varid{mult}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{};\Varid{semiring\char95 distr:>}\;\Conid{Distribute}\;\Varid{mult}\;\Varid{plus}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{};\Varid{mult\char95 0\char95 l:}\;\forall\;\Varid{x},\Varid{0}\;\Varid{*}\;\Varid{x}\;= \;\Varid{0}\mskip1.5mu\}\;.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
where e.g.\ \ensuremath{\Conid{RingPlus}} is the operation class for the semigroup operator.
The notation \ensuremath{\Conid{:>}} makes \ensuremath{\Varid{semiring\char95 mult\char95 monoid}} a coercion: when asked for a monoid it suffices to
provide a ring.

\emph{Structure-as-parameters} helps setoid-rewriting: type class resolution
can find the equivalence relation in the context.
% A similar style should be possible for, say, the |Ring| tactic, instead of
% declaring the ring structure by a separate command, we would rely on type class resolution to find
% it.
We note that \ensuremath{\Varid{op}} does not depend on the proof that \ensuremath{\Varid{e}} is an equivalence. As explained above we
use Coq's implicit quantification (\ensuremath{\hspace*{0mm}`\hspace*{0mm}\{\mskip1.5mu\}}) to avoid having to write all the parameters when
\emph{stating} a theorem and Coq's maximally inserted implicit arguments to find the parameters when
\emph{applying} a theorem. Both features are new in Coq and stem from the type class implementation.

We mention the trade-off between bigger contexts versus bigger terms. Our contexts are bigger than
those of telescopes or packed classes. In our experience, this has been relatively
harmless\footnote{Coq's data structure for contexts is not very efficient. Gonthier fears that this
may be a bottleneck for huge developments. It seems that the data structure chosen
in~\cite{asperti2009compact} will behave better.}%
: most terms in the context are there to support canonical names. Bigger terms
\emph{do} cause problems: 1. when proofs are part of mathematical objects we need to share these
proofs to allow rewriting. Moreover, it prohibits Opaque proofs and `proof irrelevance'. 2. The
projection paths may not be canonical.

@@

We have defined the category of setoids, the category of categories, the dual of a category,
products, and hence co-products, the category of $\tau$-algebras.

We provide the construction of the term-algebra for a type $\tau$ and show that it is
initial in the category of $\tau$-algebras. We proceed to prove that this provides a left-adjoint
to the forgetful functor. The resulting expression monad will be used in Section~\ref{quote} below.
